Draft -- Discriminating Seagrasses From Green Macroalgae in France and Portugal Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-05-02 docs/_tex/index.tex
adding Ana's comments

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft},
  pdfauthor={Simon Oiry; Bede Ffinian Rowe Davies; Ana I. Sousa; Philippe Rosa; Maria Laura Zoffoli; Guillaume Brunier; Pierre Gernez; Laurent Barillé},
  pdfkeywords={Drone, Remote Sensing, Seagrass, Coastal
Ecosystems, Neural Network},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Draft -- Discriminating Seagrasses From Green Macroalgae in
European Intertidal areas using high resolution multispectral drone
imagery -- Draft}
\author[1]{Simon Oiry%
\corref{cor1}%
}
 \ead{oirysimon@gmail.com} 
\author[1]{Bede Ffinian Rowe Davies%
%
}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing struggles with spatial and spectral
resolution, making it difficult to discriminate seagrass from other
macrophytes in highly heterogenous meadows. Drone images at a very high
spatial resolution offer a promising solution to address challenges
related to spatial heterogeneity and intrapixel mixture. This study
focuses on using drone acquisitions with a ten spectral band sensor
mirroring those of Sentinel-2, for mapping intertidal macrophytes and
effectively discriminating between seagrass and green macroalgae. Nine
drone flights were conducted at two different altitudes (12m and 120m)
across heterogeneous intertidal European habitats in France and
Portugal. Low altitude flights were used to train a Deep Learning
classifier based on Neural Networks to discrimintate among five
taxonomic classes of intertidal vegetation: Magnoliopsida (Seagrass),
Chlorophyceae (Green macroalgae), Phaeophyceae (Brown algae),
Rhodophyceae (Red macroalgae) and benthic Bacillariophyceae (Diatoms).
Classification of drone imagery resulted in an overall accuracy of 94\%
across all the sites and images, covering a total area of 467 000 m².
The model exhibited an accuracy of 96.4\% in identifying seagrass.
Importantly, seagrass and green algae can be discriminated, although
they share the same pigment composition. As the algorithm was developed
for a multispectral camera with ten spectral bands in the visible and
near-infrared, it could be adapted to the Multi-Spectral Instrument
(MSI) onboard Sentinel-2 thus offering promising perspectives for
satellite remote sensing of intertidal biodiversity over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between the land and oceans \citep{unsworth2022}. Seagrass meadows
provide a myriad of ecosystem services to humanity, including carbon
sequestration, oxygen production, protection against sea-level rise and
coastline erosion, and limitation of eutrophication. They serve as vital
habitats for a diverse array of marine and terrestrial species,
providing living, breeding, and feeding grounds \citetext{\citealp[
]{gardner2018}; \citealp[ ]{Zoffoli2022}; \citealp{jankowska2019}}. Due
to the concentration of human activities in coastal zones, seagrass
meadows are directly exposed to and impacted by anthropogenic pressures.
Global regression and fragmentation of seagrass meadows are currently
observed due to climate change, diseases, urbanization, land
reclamation, dredging, competition with alien species, and reduction in
water quality \citetext{\citealp[ ]{nguyen2021}; \citealp[
]{soissons2018}; \citealp[ ]{orth2006}; \citealp[ ]{lin2018}; \citealp[
]{duffy2019}; \citealp[
]{rasheed2011long}; \citealp{chefaoui2018dramatic}}. Both habitat
fragmentation and reduction, in turn, can severely compromise the
effectiveness of ecosystem services provided by seagrass meadows. While
improvements in water quality have been recently reported in Europe,
allowing an overall recovery of seagrass ecosystems at local scales,
many coastal waters worldwide are still subjected to strong
eutrophication processes \citetext{\citealp[
]{deSantos2019}; \citealp{Zoffoli2021}}. Coastal eutrophication has been
associated to excessive accumulation of green macroalgae, so-called
green tides \citep{devlin2023nutrients}. Green tides produce shade and
suffocation over seagrass individuals, thus threatening the health of
seagrass ecosystems \citep{wang2022}.

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of these
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only partially exposed during low tide and can be
situated in difficult-to-reach mudflats, potentially leading to
inaccurate and limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for the near
real-time and consistent retrieval of seagrass EOVs and EBVs over
extensive meadows. \citetext{\citealp[ ]{Zoffoli2021}; \citealp[
]{xu2021}; \citealp[ ]{Traganos2018}; \citealp{coffer2023}}

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However the application of this
approach to satellite RS remains to be validated. Moreover patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones can potentially fill the data gaps left by satellite RS and
\emph{in situ} measurements, due to their ability to provide
spatially-explicit observations at very high spatial resolutions (pixel
size from mm to cm) while capturing data at multi-spectral resolution
\citetext{\citealp[ ]{fairley2022drone}; \citealp{oh2017use}}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management \citetext{\citealp[
]{adade2021}; \citealp[ ]{casella2020}; \citealp{angnuureng2022}} to
mapping species distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[
]{Roman2021}; \citealp{Brunier2022Topographic}}. However, when applied
to coastal habitat mapping, most case studies were previously limited to
a single flight, restricting the generalizability of their application
over wider geographical scales \citetext{\citealp[
]{Roman2021}; \citealp[ ]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes over a diverse biogeographical range, with a
particular focus on discriminating Magnoliopsida and Chlorophyceae
(Seagrass and Green Algae respectively). Ten drone flights were
performed over soft-bottom intertidal areas along the Atlantic
coastlines of two European countries (France and Portugal), covering a
wide range of habitats, from monospecific seagrass meadows to meadows
mixed with green, or red macroalgae. A deep learning algorithm was
trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy.

@@

Seven study sites distributed between France and Portugal were selected
for their relatively extensive intertidal seagrass beds. Two sites were
located in the Gulf of Morbihan (Figure~\ref{fig-map} A), France
(47.5791°N, 2.8018°W). This gulf covers an area of 115 km² and is only
connected to the sea through a 900m wide channel. A total of 53 small
islands are scattered across the gulf leading to 250 km of shorelines.
Patchy seagrass meadows can be found on many of these islands. One of
the sites within the gulf was on one its islands (Arz) and the other was
located further south on a mainland beach area (Duer). Two other sites
were located in Bourgneuf Bay, France (46.9849°N, 2.1488°W). This is a
340 km² semi-enclosed macrotidal bay, protected from waves by
Noirmoutier Island. Bourgneuf bay hosts a large intertidal seagrass
meadow of about 6 km² \citep{ZOFFOLI2020112020}. Within this meadow, the
sites observed by drones (L'Epine and Barbatre, Figure~\ref{fig-map} B)
contained monospecific beds of \emph{Nanozostera noltei} (dwarf
eelgrass) with very little mixing with other macrophytes. Three sites
were surveyed in the Ria de Aveiro Lagoon in Portugal (40.6887°N,
8.6810°W). The extent of this lagoon is \textasciitilde80 km² with many
narrow channels, large salt marshes and many mudflats that uncover at
low tide. It is connected to the open sea through a single channel, with
a tidal lag between the North and the South of the lagoon. The
southernmost site (Gafanha) was a mudflat located in a tributary river
of the lagoon whereas the two other sites (Mataducos and Marinha
Lanzarote) were situated in the middle of the lagoon and only accessible
by boat (Figure~\ref{fig-map} C). These Portuguese sites are
characterized by a more diverse intertidal vegetation, where patches of
seagrass intermingle with red, brown, and green macroalgae.

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(Unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
seagrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Pictures of each quadrat were
uploaded online to the Global Biodiversity Information Facility (GBIF)
platform \citep{BedeGbif}, a public open portal to store and share
biodiversity data. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to error reduction in photo interpretation.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and their raw spectral signatures
at the spectral resolution of the Micasense RedEdge Dual MX.}

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft}was used to process images to obtain multispectral
orthomosaics of each flight. The process for orthomosaicking was
identical for every flight. First, tying key points were detected inside
of each image and between overlapping images in order to obtain a sparse
point cloud. This cloud was cleaned using reprojection accuracy metric
in order to remove noisy points. A dense point cloud was then produced
using a structure from motion algorithm. A surface interpolation of this
dense point cloud was made to obtain a digital surface model (DSM), used
to reconstruct the multispectral ortho-image \citep{nebel2020review}.
Low altitude drone flights produced ortho-images with a very high
spatial resolution (8 mm per pixel), making it efficient to visually
distinguish between the various types of vegetation. High altitude
fights on the other hand allowed to cover large areas and produced
images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

\subsection{Impact of the percent cover on the
prediction}\label{impact-of-the-percent-cover-on-the-prediction}

The key aspect of the workflow adopted in the present study is the
mapping of specific areas at two different altitudes (12m and 120m),
resulting in two distinct resolutions for the same area (8mm and 80mm).
The predictions made on the high-resolution flight can be used to
estimate the percent cover of each vegetation class for every pixel of
the lower resolution flight. Consequently, for each pixel of the
high-altitude flights, percent covers of each vegetation class were
obtained, and a kernel density plot was generated. This plot provides a
visual representation of the behavior of the model in different
vegetation percent cover scenarios.

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model derived likelihood of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represents the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with
Chlorophyceae and Rhodophyceae mixed with the seagrass meadow. There was
also Bacillariophyceae forming biofilms on parts of sediment surface.
Although the seagrass bed was solely composed of \emph{Nanozostera
noltei}, various colors of this specie could be observed from dark green
(corresponding to healthy leaves) to whitish/brown (when leaves were
discolored having lost their pigmentation). Regardless of color,
Magnoliopsida pixels were accurately predicted by the model.

@@

\caption{\label{fig-GafLow}RGB orthomosaic (Left) and Prediction (Right)
of the low altitude flight of Gafanha, Portugal. The total extent of
this flight was 3000m² with a resolution of 8 mm per pixel. Background
colors indicate intertidal area (Light Green) and land area (Light
Grey). The zoom covers an area equivalent to a 10-meter Sentinel-2 pixel
size.}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Spartina sp.} were
misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

Amoung the high altitude flights, the one acquired over the inner lagoon
of the Ria de Aveiro covered the largest area with approximately 1.5 km²
(Figure~\ref{fig-Boat}). This site was dominated by seagrass and red
macroalgae. The classification provided consistent results, with a
patchy seagrass meadow mixed with red macroalgae on the eastern part of
the site. As shown in the zoom (Figure~\ref{fig-Boat}), the edges of the
meadow visually appears to be colonised by green macroalgae, which the
model agreed with.

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12m and 120m) for constructing the neural
network model. Overall there was a good agreement between the two
altitudes (Figure~\ref{fig-CompareRef}).

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The longer the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardised reflectance spectra of each class.
Each slice represents the VIP of both raw and standardised reflectance
combined.}

@@

When the percent cover of a given class was 100 \%, coarser high-flight
pixels were well classified for all the classes except for Bare
Sediment, which was only well classified 80\% of the time. This
phenomenon may be attributed to the time gap between the two flights,
allowing for microphytobenthos migration to the surface during low tide,
consequently altering the model's classification from bare sediment to
Bacillariophyceae. A percent cover of at least 80\% was sufficient to
have all the pixels of high altitude flights correctly classified, with
the exception of Magnoliopsida that required a higher percent cover
(\textgreater90 \%) to be well classified. Concerning the probability of
each class, the highest percent cover was needed to confidently predict
Bacillariophyceae. To predict Chlorophyceae with a model likelihood of
0.85, a percent cover of 93 \% was needed, 90 \% for Magnoliopsida, 92
\% for Rhodophyceae and 97 \% for Bacillariophyceae.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Draft -- Discriminating Seagrasses From Green Macroalgae in France and Portugal Intertidal areas using high resolution multispectral drone imagery -- Draft},
  pdfauthor={Simon Oiry; Bede Ffinian Rowe Davies; Ana I. Sousa; Philippe Rosa; Maria Laura Zoffoli; Guillaume Brunier; Pierre Gernez; Laurent Barillé},
  pdfkeywords={Drone, Remote Sensing, Seagrass, Coastal
Ecosystems, Neural Network},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Draft -- Discriminating Seagrasses From Green Macroalgae in
France and Portugal Intertidal areas using high resolution multispectral
drone imagery -- Draft}
\author[1]{Simon Oiry%
\corref{cor1}%
}
 \ead{oirysimon@gmail.com} 
\author[1]{Bede Ffinian Rowe Davies%
%
}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing struggles with spatial and spectral
resolution, making it difficult to discriminate seagrass from other
macrophytes in highly heterogenous meadows. Drone (unmanned aerial
vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor mirroring those of
Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12 m and 120 m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. Importantly, seagrass and green algae
can be discriminated, although they share the same pigment composition.
As the algorithm was developed for a multispectral camera with ten
spectral bands in the visible and near-infrared, it could be adapted to
the Multi-Spectral Instrument (MSI) onboard Sentinel-2 thus offering
promising perspectives for satellite remote sensing of intertidal
biodiversity over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between the land and oceans \citep{unsworth2022}. Seagrass meadows
provide a myriad of ecosystem services to humanity, including carbon
sequestration, oxygen production, protection against sea-level rise and
coastline erosion, and mitigation of eutrophication \citetext{\citealp[
]{unsworth2022}; \citealp{sousa2019blue}}. They serve as vital habitats
for a diverse array of marine and terrestrial species, providing living,
breeding, and feeding grounds \citetext{\citealp[
]{gardner2018}; \citealp[ ]{Zoffoli2022}; \citealp{jankowska2019}}. Due
to the concentration of human activities in coastal zones, seagrass
meadows are directly exposed to and impacted by anthropogenic pressures.
Global regression and fragmentation of seagrass meadows are currently
observed due to climate change, diseases, urbanization, land
reclamation, dredging, competition with alien species, and reduction in
water quality \citetext{\citealp[ ]{nguyen2021}; \citealp[
]{soissons2018}; \citealp[ ]{orth2006}; \citealp[ ]{lin2018}; \citealp[
]{duffy2019}; \citealp[ ]{rasheed2011long}; \citealp[
]{chefaoui2018dramatic}; \citealp{sousa2019blue}}. Both habitat
fragmentation and reduction, in turn, can severely compromise the
effectiveness of ecosystem services provided by seagrass meadows. While
improvements in water quality and hydrodynamics have been recently
reported in Europe, allowing an overall recovery of seagrass ecosystems
at local and European scales, many coastal waters worldwide are still
subjected to strong eutrophication processes \citetext{\citealp[
]{deSantos2019}; \citealp[ ]{Zoffoli2021}; \citealp{sousa2019blue}}.
Coastal eutrophication has been associated to excessive accumulation of
green macroalgae, so-called green tides \citep{devlin2023nutrients}.
Green tides produce shade and suffocation over seagrass individuals,
thus threatening the health of seagrass ecosystems \citep{wang2022}.

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of these
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only exposed during low tide and can be situated in
difficult-to-reach mudflats, potentially leading to inaccurate and
limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for the near
real-time and consistent retrieval of seagrass EOVs and EBVs over
extensive meadows. \citetext{\citealp[ ]{Zoffoli2021}; \citealp[
]{xu2021}; \citealp[ ]{Traganos2018}; \citealp{coffer2023}}

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100 m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10 m) or Landsat8/9 (30 m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition \citetext{\citealp[
]{ralph2002}; \citealp{Douay2022}}. Therefore, to someone not
specialized in the field, their spectral signatures may appear to be
alike \citetext{\citealp[ ]{Davies2023}; \citealp{bannari2022}}.
Recently, using advanced machine-learning algorithms trained with a
large hyperspectral library of more than 300 field reflectance spectra,
\citep{Davies2023} demonstrated that it was possible to discriminate
Magnoliopsida from Chlorophyceae using radiometric data acquired at
Sentinel-2 's spectral resolution. However the application of this
approach to satellite RS remains to be validated. Moreover patches of
green algae can develop at small spatial scales that are not observable
using non-commercial satellite imagery \citep{tuya2013}, especially
during the initial stage of a green tide.

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citetext{\citealp[
]{fairley2022drone}; \citealp{oh2017use}}. The versatility of drones
allows for their application across a diverse thematic range , from
coastal zone management \citetext{\citealp[ ]{adade2021}; \citealp[
]{casella2020}; \citealp{angnuureng2022}} to mapping species
distribution \citetext{\citealp[ ]{joyce2023}; \citealp[
]{tallam2023}; \citealp[ ]{Roca2022}; \citealp[ ]{Roman2021}; \citealp[
]{Brunier2022Topographic}; \citealp{sousa2019blue}}. However, when
applied to coastal habitat mapping, most case studies were previously
limited to a single flight, restricting the generalizability of their
application over wider geographical scales \citetext{\citealp[
]{Roman2021}; \citealp[ ]{collin2019improving}; \citealp[
]{rossiter2020uav}; \citealp{Brunier2022Topographic}}. The present study
aimed at analyzing the potential of multispectral drone RS to map
intertidal macrophytes over a diverse biogeographical range, with a
particular focus on discriminating Magnoliopsida and Chlorophyceae
(Seagrass and Green Algae, respectively). Ten drone flights were
performed over soft-bottom intertidal areas along the Atlantic
coastlines of two European countries (France and Portugal), covering a
wide range of habitats, from monospecific seagrass meadows to meadows
mixed with green, or red macroalgae. A deep learning algorithm was
trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy.

@@

Seven study sites distributed between France and Portugal were selected
for their relatively extensive intertidal seagrass beds. Two sites were
located in the Gulf of Morbihan (Figure~\ref{fig-map} A), France
(47.5791°N, 2.8018°W). This gulf covers an area of 115 km² and is only
connected to the sea through a 900 m wide channel. A total of 53 small
islands are scattered across the gulf leading to 250 km of shorelines.
Patchy seagrass meadows can be found on many of these islands. One of
the sites within the gulf was on one its islands (Arz) and the other was
located further south on a mainland beach area (Duer). Two other sites
were located in Bourgneuf Bay, France (46.9849°N, 2.1488°W). This is a
340 km² semi-enclosed macrotidal bay, protected from waves by
Noirmoutier Island. Bourgneuf bay hosts a large intertidal seagrass
meadow of about 6 km² \citep{ZOFFOLI2020112020}. Within this meadow, the
sites observed by drones (L'Epine and Barbatre, Figure~\ref{fig-map} B)
contained monospecific beds of \emph{Nanozostera noltei} (dwarf
eelgrass, syn. \emph{Zostera noltei}) with very little mixing with other
macrophytes. Three sites were surveyed in the Ria de Aveiro Coastal
Lagoon in Portugal (40.6887°N, 8.6810°W). The extent of this lagoon is
\textasciitilde83 km² (at low tide) with many narrow channels, large
salt marshes and many mudflats that uncover at low tide
\citep{sousa2017blue}. It is connected to the open sea through a single
channel, with a tidal lag between the North and the South of the lagoon.
The southernmost site (Gafanha) is a mudflat located in the Mira channel
(one of the four main channels of the lagoon) whereas the two other
sites (Mataducos and Marinha Lanzarote) were situated in the middle of
the lagoon and only accessible by boat (Figure~\ref{fig-map} C). These
Portuguese sites are characterized by a more diverse intertidal
vegetation, where patches of seagrass intermingle with red, brown, and
green macroalgae.

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
eelgrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Pictures of each quadrat were
uploaded online to the Global Biodiversity Information Facility (GBIF)
platform \citep{BedeGbif}, a public open portal to store and share
biodiversity data. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to error reduction in photo interpretations.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and their raw spectral signatures
at the spectral resolution of the Micasense RedEdge Dual MX. A :
Magnoliopsida (Nanozostera noltei syn. Zostera noltei) ; B :
Phaeophyceae (Fucus sp.) ; C : Rhodophyceae (Gracilaria vermiculophylla)
; D : Chlorophyceae (Ulva sp.) ; E : Bacillariophyceae (Diatoms - MPB)}

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft}was used to process images to obtain multispectral
orthomosaics of each flight. The process for orthomosaicking was
identical for every flight. First, tying key points were detected inside
of each image and between overlapping images in order to obtain a sparse
point cloud. This cloud was cleaned using reprojection accuracy metric
in order to remove noisy points. A dense point cloud was then produced
using a structure from motion algorithm. A surface interpolation of this
dense point cloud was made to obtain a digital surface model (DSM), used
to reconstruct the multispectral ortho-image \citep{nebel2020review}.
Low altitude drone flights produced ortho-images with a very high
spatial resolution (8 mm per pixel), making it efficient to visually
distinguish between the various types of vegetation. High altitude
flights on the other hand allowed to cover large areas and produced
images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

\subsection{Impact of the vegetation cover on the
prediction}\label{impact-of-the-vegetation-cover-on-the-prediction}

The key aspect of the workflow adopted in the present study is the
mapping of specific areas at two different altitudes (12m and 120m),
resulting in two distinct resolutions for the same area (8mm and 80mm).
The predictions made on the high-resolution flight can be used to
estimate the cover (percentage, \%) of each vegetation class for every
pixel of the lower resolution flight. Consequently, for each pixel of
the high-altitude flights, the cover (\%) of each vegetation class were
obtained, and a kernel density plot was generated. This plot provides a
visual representation of the behavior of the model in different
vegetation percent cover scenarios.

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model derived likelihood of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represents the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with
Chlorophyceae and Rhodophyceae mixed with the seagrass meadow. There was
also Bacillariophyceae forming biofilms on parts of bare sediment
surface. Although the seagrass bed was solely composed of
\emph{Nanozostera noltei}, various colors of this specie could be
observed from dark green (corresponding to healthy leaves) to
whitish/brown (when leaves were discolored having lost their
pigmentation). Regardless of color, Magnoliopsida pixels were accurately
predicted by the model.

@@

\caption{\label{fig-GafLow}RGB orthomosaic (Left) and Prediction (Right)
of the low altitude flight of Gafanha, Portugal. The total extent of
this flight was 3000 m² with a resolution of 8 mm per pixel. Background
colors indicate intertidal area (Light Green) and land area (Light
Grey). The zoom covers an area equivalent to a 10-meter Sentinel-2 pixel
size.}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the intertidal
area was classified as Magnoliopsida by the model, including seagrass
patches with discolored leaves. Only a few pixels were classified as
Chlorophyceae at this spatial scale. Furthermore, the area that was
classified as Bacillariophyceae in the low-altitude flight remained
mostly classified as such in the high-altitude flight, though some
pixels were classified as Magnoliopsida. Patches of Rhodophyceae were
correctly classified. In the northern part of the site and near the land
eadges, patches of the schorre angiosperm \emph{Sporobolus maritimus}
(syn. \emph{Spartina maritima)} were misclassified, either as
Magnoliopsida or as Phaeophyceae.

@@

Amoung the high altitude flights, the one acquired over the inner part
of Ria de Aveiro coastal lagoon covered the largest area with
approximately 1.5 km² (Figure~\ref{fig-Boat}). This site was dominated
by seagrass and red macroalgae. The classification provided consistent
results, with a patchy seagrass meadow mixed with red macroalgae on the
eastern part of the site. As shown in the zoom (Figure~\ref{fig-Boat}),
the edges of the meadow visually appears to be colonised by green
macroalgae (\emph{Ulva sp.}), which the model agreed with.

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12 m and 120 m) for constructing the neural
network model. Overall there was a good agreement between the two
altitudes (Figure~\ref{fig-CompareRef}).

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The longer the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardised reflectance spectra of each class.
Each slice represents the Variable Importance (VI) of both raw and
standardised reflectance combined.}

@@

When the vegetation cover of a given class was 100 \%, coarser
high-flight pixels were well classified for all the classes except for
Bare Sediment, which was only well classified 80\% of the time. This
phenomenon may be attributed to the time gap between the two flights,
allowing for microphytobenthos migration to the surface during low tide,
consequently altering the model's classification from bare sediment to
Bacillariophyceae. A percent cover of at least 80\% was sufficient to
have all the pixels of high altitude flights correctly classified, with
the exception of Magnoliopsida that required a higher cover
(\textgreater90 \%) to be well classified. Concerning the probability of
each class, the highest percent cover was needed to confidently predict
Bacillariophyceae. To predict Chlorophyceae with a model likelihood of
0.85, a cover of 93 \% was needed, 90 \% for Magnoliopsida, 92 \% for
Rhodophyceae and 97 \% for Bacillariophyceae.
